High-Dimensional Regularized Discriminant Analysis

### 2016-12-02 reports/paper-hdrda.Rnw
Describe timing comparisons

\author[jar]{John A. Ramey\cortext[cor1]{Corresponding author}}
\author[cks]{Caleb K. Stein}
\author[pdy]{Phil D. Young}
\author[dmy]{Dean M. Young}
\address[jar]{uStudio Labs}
\address[cks]{Myeloma Institute, University of Arkansas for Medical Sciences}
\address[pdy]{Department of Management and Information Systems, Baylor University}
\address[dmy]{Department of Statistical Science, Baylor University}

@@

In this section, we demonstrate that the computational performance of the
training and model selection employed in the \emph{HDRDA} classifier is
substantially faster than that of the \emph{RDA} classifier on small-sample,
high-dimensional data sets. The relative difference in runtime between the two
classifiers drastically increases as the number of features increases. To
compare the two classifiers, we generate 25 observations from each of $K=4$
multivariate normal populations with mean vectors $\bm \mu_1 = -3 \cdot \bm
1_p$, $\bm \mu_2 = -\bm 1_p$, $\bm \mu_3 = \bm 1_p$, and $\bm \mu_4 = 3 \cdot
\bm 1_p$. We set the covariance matrix of each population to the $p \times p$
identity matrix. For each data set generated, we estimate the $\lambda$ and
$\gamma$ parameters for both classifiers using a grid of 21 equidistant
candidate values between 0 and 1, inclusively. We set $\alpha_k = 1 - \gamma$,
$k = 1, \ldots, K$, in the \emph{HDRDA} classifier to resemble
\citeauthor{Friedman:1989tm}'s \emph{RDA} parameterization. At each pair of
$\lambda$ and $\gamma$, we compute the $V$-fold cross-validation error rate with
$V = 10$ \citep{Hastie:2008dt}. The optimal model is the pair of $\lambda$ and
$\gamma$ that minimizes the $V$-fold cross-validation error rate.

Here, our goal is to compare the computational runtime of the model selection
for both classifiers. We compare the runtime of both classifiers by increasing
the number of features from $p = 500$ to $p = 5000$ in increments of 500. We
generate 100 data sets for each value of $p$ and compute the training and model
selection runtime of both classifiers. Our timing comparisons are based on our
\emph{HDRDA} implementation in the {\tt sparsediscrim} R package with the
standard \emph{RDA} implementation in the {\tt klaR} R package. All timing
comparisons were conducted on an Amazon Elastic Compute Cloud (EC2) {\tt
  c4.4xlarge} instance using version 3.3.1 of the open-source statistical
software {\tt R}. Our analyses can be reproduced using the code available in the
\url{https://github.com/ramhiser/paper-hdrda} repository.

@@

TODO
\[ \left[\text{Insert Figure \ref{fig:timing-results} approximately here }\right] \]

@@

The \emph{HDRDA} implementation of \eqref{eq:hdrda-decomposed} in the {\tt
  sparsediscrim} R package was comparable to that of the diagonal classifiers in
terms of computational runtime. Given the \emph{RDA} classifier from
\cite{Friedman:1989tm} has been shown to have excellent performance in the
high-dimensional setting \citep{Webb:2011vu} but is limited by its
computationally intense model-selection procedure, our work enables the
\emph{RDA} classifier to be employed with small-sample, high-dimensional data in
practice. This result is reassuring because the \emph{RDA} classifier remains
widely popular in the literature. In fact, variants of the RDA classifier have
been applied to microarray data
\citep{Ching:2012fu,Li:2012ev,Tai:2007bk,Guo:2007te}, facial recognition
\citep{Ching:2012fu,Zhang:2010va,Dai:2007vd,Lu:2005hq,Pima:2004vw,Lu:2003we},
handwritten digit recognition \citep{Bouveyron:2007gx}, remote sensing
\citep{Tadjudin:1999fk}, seismic detection \citep{Anderson:2002kg}, and chemical
spectra \citep*{Wu:1996us,Aeberhard:1993fp}.

>>>>

\author[jar]{John A. Ramey\cortext[cor1]{Corresponding author}}
\author[cks]{Caleb K. Stein}
\author[pdy]{Phil D. Young}
\author[dmy]{Dean M. Young}
\address[jar]{Novi Labs}
\address[cks]{Myeloma Institute, University of Arkansas for Medical Sciences}
\address[pdy]{Department of Management and Information Systems, Baylor University}
\address[dmy]{Department of Statistical Science, Baylor University}

@@

In this section, we demonstrate that the computational performance of the
training and model selection employed in the \emph{HDRDA} classifier is
substantially faster than that of the \emph{RDA} classifier on small-sample,
high-dimensional data sets. The relative difference in runtime between the two
classifiers drastically increases as the number of features increases. To
compare the two classifiers, we generate 25 observations from each of $K=4$
multivariate normal populations with mean vectors $\bm \mu_1 = -3 \cdot \bm
1_p$, $\bm \mu_2 = -\bm 1_p$, $\bm \mu_3 = \bm 1_p$, and $\bm \mu_4 = 3 \cdot
\bm 1_p$. We set the covariance matrix of each population to the $p \times p$
identity matrix. For each data set generated, we estimate the $\lambda$ and
$\gamma$ parameters for both classifiers using a grid of 5 equidistant candidate
values between 0 and 1, inclusively. We set $\alpha_k = 1 - \gamma$, $k = 1,
\ldots, K$, in the \emph{HDRDA} classifier to resemble
\citeauthor{Friedman:1989tm}'s \emph{RDA} parameterization. At each pair of
$\lambda$ and $\gamma$, we compute the 10-fold cross-validation error rate
\citep{Hastie:2008dt}. Then, we select the optimal model that minimizes the 10-fold
cross-validation error rate.

Here, our goal is to compare the computational runtime of the model selection
for both classifiers as a function of the number of features $p$. We compare the
runtime of both classifiers by increasing the number of features from $p = 500$
to $p = 5000$ in increments of 500. We generate 100 data sets for each value of
$p$ and compute the training and model selection runtime of both
classifiers. Our timing comparisons are based on our \emph{HDRDA} implementation
in the {\tt sparsediscrim} R package and the standard \emph{RDA} implementation
in the {\tt klaR} R package. All timing comparisons were conducted on an Amazon
Elastic Compute Cloud (EC2) {\tt c4.4xlarge} instance using version 3.3.1 of the
open-source statistical software {\tt R}. Our timing comparisons can be
reproduced with the code available in the
\url{https://github.com/ramhiser/paper-hdrda} repository.

@@

In Figure \ref{fig:timing-results}, we plot the runtime of the model selections
for both the \emph{HDRDA} and \emph{RDA} classifiers as a function of the number
of features $p$. In the left panel of Figure \ref{fig:timing-results}, we
observe that the \emph{HDRDA} classifier is substantially faster than the
\emph{RDA} classifier as $p$ increases. Figure \ref{fig:timing-results} suggests
that the usage of a matrix inverse and determinant in the {\tt klaR} R package's
discriminant function yielded training and model-selection timings that exceeded
linear growth in $p$. Because our implementation in the \emph{HDRDA} classifier
removes the inverse and determinants from the classifier, the \emph{HDRDA}
classifier is computationally more efficient than the \emph{RDA} classifier,
especially as $p$ increases. The variance in the \emph{RDA} classifier's runtime
also appeared to increase as $p$ increased. To quantify how much beyond linear
growth in $p$, we fit a regression line quadratic in $p$ to the \emph{RDA}
runtimes and plotted the fit in Figure \ref{fig:timing-results}. We attempted to
fit a cubic polynomial in $p$ to the \emph{RDA} timings, but the cubic term was
insignificant with significant quadratic and linear terms. We also fit two
polynomials to the \emph{RDA} timings with orders 2.807 and 2.376, which
correspond to the asymptotic computational runtime of two well-known algorithms
for computing a matrix inverse: the Strassen and Coppersmith-Winograd
algorithms, respectively \footnote{TODO: Citations needed}. In both cases, the
polynomials were statistically significiant at the 0.05 level. We displayed the
quadratic fit in Figure \ref{fig:timing-results} because it fit the timings best
for $p = 500$ to $p = 2000$. The fits for $2000 \le p \le 5000$ were similar.

In the left panel of Figure \ref{fig:timing-results}, we also fit a linear
regression model that was linear in $p$ to the \emph{HDRDA} runtimes. In the
right panel, we repeated the same scatterplot and linear fit with the timings
restricted to the observed range of \emph{HDRDA} timings. The outliers in the
\emph{HDRDA} timings were primarily due to the inherent computing variability in
our {\tt c4.4xlarge} instance.

Clearly, the model training and selection employed in \emph{HDRDA} is
substantially faster than that of \emph{RDA}. To investigate the relative timing
comparisons between the two classifiers, we consider the ratio of mean timings
of the \emph{RDA} classifier to the \emph{HDRDA} classifier for each value of
$p$. We employed nonparametric bootstrapping\footnote{TODO: Citation needed} to
estimate the mean ratio along with 95\% confidence intervals. In Figure
\ref{fig:timing-comparison-bootstrap}, the bootstrap sampling distributions for
the ratio of mean timings are given. First, we observe that the mean relative
timings increases as $p$ increases. For smaller dimensions, the relative
difference in computing is sizable with the average ratio of the mean timings
equal to \Sexpr{bootstrap_summary_p500$avg_ratio} for $p = 500$ and a 95\%
confidence interval of (\Sexpr{bootstrap_summary_p500$lower_ci},
\Sexpr{bootstrap_summary_p500$upper_ci}). Furthermore, the ratio of mean
computing times suggests that the \emph{RDA} classifier is impractical for
higher dimensions. For instance, when $p = 5000$ the ratio of mean computing
times increases to \Sexpr{bootstrap_summary_p5000$avg_ratio} with a 95\%
confidence interval of (\Sexpr{bootstrap_summary_p5000$lower_ci},
\Sexpr{bootstrap_summary_p5000$upper_ci}).

\[ \left[\text{Insert Figure \ref{fig:timing-results} approximately here }\right] \]

@@

We demonstrated that the \emph{HDRDA} implementation of
\eqref{eq:hdrda-decomposed} in the {\tt sparsediscrim} R package can be used in
practice with high-dimensional data sets. For instance, in our timing
comparisons we demonstrated that training and model selection could be employed
on data sets having 5,000 features in 3 seconds on average. We contrast our
implementation to the the poor computational performance of the in the {\tt
  klaR} R package, which took approximately 25 minutes on average to train the
\emph{RDA} classifier on a data set having 5,000 features. Given the \emph{RDA}
classifier has been shown to have excellent performance in the high-dimensional
setting \citep{Webb:2011vu} but is limited by its computationally intense
model-selection procedure, our work enables the \emph{RDA} classifier to be
employed with small-sample, high-dimensional data in practice. This result is
reassuring because the \emph{RDA} classifier remains widely popular in the
literature. In fact, variants of the RDA classifier have been applied to
microarray data \citep{Ching:2012fu,Li:2012ev,Tai:2007bk,Guo:2007te}, facial
recognition
\citep{Ching:2012fu,Zhang:2010va,Dai:2007vd,Lu:2005hq,Pima:2004vw,Lu:2003we},
handwritten digit recognition \citep{Bouveyron:2007gx}, remote sensing
\citep{Tadjudin:1999fk}, seismic detection \citep{Anderson:2002kg}, and chemical
spectra \citep*{Wu:1996us,Aeberhard:1993fp}.
